msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design

### 2025-06-17 paper/msqrob2tmt_rebuttal.md

## Reviewer 3

We thank reviewer 3 for their positive response. There has indeed been a decrease in image resolution during the auto-preparation, but this does not match with the image resolution of the figures we submitted. We will make sure that high resolution images are included in the publication before accepting the final proof.

## Reviewer 4

We thank reviewer 4 for their comments. Below, we provide an answer to each concern and indicate how we adapted the manuscript accordingly.

@@

Next, we would like to point out that the FDP is the false discovery proportion i.e., the fraction of false positives in the list of returned DA proteins, which can typically not be evaluated for real experiments as the ground truth is unknown. Therefore, the FDR has been developed, which is the expected FDP i.e., the average of the FDPs that would be obtained when the experiment were to be repeated an infinite number of times. In this contribution we estimate the FDR with the Benjamini Hochberg method, which implies a number of assumptions, e.g. independence between features (proteins), a known distribution of the test statistic under the null hypothesis and thus assumptions on the distribution of the data, etc.

Interestingly, the FDP can be computed for the spike-in studies that are used to benchmark tools as the spike-in proteins are known to be DA and the background proteins are not DA, which we exploit in our contribution to construct performance curves (sensitivity vs FDP curves). In order to evaluate the FDR control of the different workflows, we would need many repeats of the spike-in study so as to assess if the average FDP equals the nominal FDR-threshold that was used.

As there are no repeats available for the spike-in studies, we have opted to plot the observed FDP at the nominal 5% FDR-level on the performance curves as an indication on how well a workflow can control the FDR.

Note, that the FDP in a specific spike-in study is expected to deviate from the nominal 5% FDR-level by random chance, as the FDR corresponds to the FDP we expect on average rather than the FDP for the single spike-in study.

### 2025-06-17 paper/manuscript_tc.qmd
Updated latest version upon peer-review

---
title: "msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design"
authors:
  - name: Stijn Vandenbulcke
    affiliations:
      - ref: VIB
      - ref: Biomed
      - ref: UGent
    equal-contributor: true
  - name: Christophe Vanderaa
    affiliations:
      - ref: UGent
    equal-contributor: true
  - name: Oliver Crook
    affiliations:
      - ref: UK
  - name: Lennart Martens
    affiliations:
      - ref: VIB
      - ref: Biomed
  - name: Lieven Clement
    affiliations:
      - ref: UGent
    email: lieven.clement@ugent.be
    corresponding: true

affiliations:
  - id: VIB
    name: VIB-UGent Center for Medical Biotechnology, VIB, Ghent, Belgium
  - id: Biomed
    name: Department of Biomolecular Medicine, Ghent University, Ghent, Belgium
  - id: UGent
    name: Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium
  - id: UK
    name: Department of Statistics, University of Oxford, Oxford, UK.

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has the following design: \sout{UPS1 peptides at concentrations of 500, 333, 250,  and 62.5 fmol}\textcolor{red}{500, 333, 250, and 62.5 fmol of UPS1 peptides} were spiked into 50 µg of SILAC HeLa peptides. This series forms a dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest UPS1 peptide \sout{concentration}\textcolor{red}{amount} (500 fmol). A reference sample was prepared by combining the diluted UPS1 peptide samples \textcolor{red}{(286.5 fmol)} with 50 µg of SILAC HeLa peptides. Each dilution and the reference sample were processed in duplicate, yielding a total of ten samples. These samples were subsequently labelled using TMT10-plex reagents and combined for LC-MS/MS analysis, hereafter referred to as a "mixture." This protocol was repeated five times. Each mixture was analysed in technical triplicate, resulting in a total of 15 MS runs. This experimental design simulates a scenario with 10 biological replicates per condition, consisting of 2 replicates per condition within each of the five mixtures. \textcolor{red}{The MS data were searched by the authors using Proteome Discoverer 2.2.0.388 (Thermo Fisher Scientific) and Mascot Server 2.6.1 (Matrix Science, London, UK).}

@@

These findings are further corroborated by the results at the 5% FDR level, shown in @fig-WorkflowComparisons C. All msqrob2 workflows demonstrate high sensitivity and low FDP. Specifically, they recover between `r rangeMsqrobTP` \textcolor{red}{DA hits (true positives,TP) for} spike-in UPS proteins \sout{as DA (true positives,TP)}\textcolor{red}{across all 6 pairwise comparisons} while only reporting between `r rangeMsqrobFP` false positives (FP) for HeLa proteins. As a result, their FDP ranges between `r rangeMsqrobFDP`%, suggesting appropriate FDR control at the 5% level. The default msTrawler workflow, however, was only able to recover `r msTrawlerTP` TP and reported `r msTrawlerFP` FP, leading to an FDP of `r msTrawlerFDP`%. With our refactored import function this improved to `r msTrawlerTP_fixed` TP, `r msTrawlerFP_fixed` FP and an FDP of `r msTrawlerFDP_fixed`%. The summarisation-based workflows DEqMS and MSstatsTMT, reported `r deqmsTP` and  `r msStatsTmtTP`  TP, respectively, with `r deqmsFP` and `r msStatsTmtFP` FP, resulting in FDPs of `r deqmsFDP`% and `r msStatsTmtFDP`%, respectively, suggesting improper FDR control by DEqMS.

@@

\section{Data and Code Availability}

>>>>

---
title: "msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design"
authors:
  - name: Stijn Vandenbulcke
    affiliations:
      - ref: UGent
      - ref: VIB
      - ref: Biomed
    equal-contributor: true
  - name: Christophe Vanderaa
    affiliations:
      - ref: UGent
    equal-contributor: true
  - name: Oliver Crook
    affiliations:
      - ref: UK
  - name: Lennart Martens
    affiliations:
      - ref: VIB
      - ref: Biomed
  - name: Lieven Clement
    affiliations:
      - ref: UGent
    email: lieven.clement@ugent.be
    corresponding: true

affiliations:
  - id: UGent
    name: Department of Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium
  - id: VIB
    name: CompOmics, VIB Center for Medical Biotechnology, VIB, Ghent, Belgium
  - id: Biomed
    name: Department of Biomolecular Medicine, Faculty of Medicine and Health Sciences, Ghent University, Ghent, Belgium
  - id: UK
    name: Department of Statistics, University of Oxford, Oxford, UK.

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has the following design: \sout{UPS1 peptides at concentrations of 500, 333, 250,  and 62.5 fmol}\textcolor{red}{500, 333, 250, and 62.5 fmol of UPS1 peptides} were spiked into 50 µg of SILAC HeLa peptides. This series forms a dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest UPS1 peptide \sout{concentration}\textcolor{red}{amount} (500 fmol). A reference sample was prepared by combining the diluted UPS1 peptide samples \textcolor{red}{(286.5 fmol)} with 50 µg of SILAC HeLa peptides. Each dilution and the reference sample were processed in duplicate, yielding a total of ten samples. These samples were subsequently labelled using TMT10-plex reagents and combined for LC-MS/MS analysis, hereafter referred to as a "mixture." This protocol was repeated five times. Each mixture was analysed in technical triplicate, resulting in a total of 15 MS runs. This experimental design simulates a scenario with 10 biological replicates per condition, consisting of 2 replicates per condition within each of the five mixtures. \textcolor{red}{The MS data were searched by the MSstatsTMT authors using Proteome Discoverer 2.2.0.388 (Thermo Fisher Scientific) and Mascot Server 2.6.1 (Matrix Science, London, UK).}

@@

These findings are further corroborated by the results at the 5% FDR level, shown in @fig-WorkflowComparisons C. All msqrob2 workflows demonstrate high sensitivity and low FDP. Specifically, they recover between `r rangeMsqrobTP` \textcolor{red}{DA hits (true positives, TP) for} spike-in UPS proteins \sout{as DA (true positives, TP)}\textcolor{red}{across all 6 pairwise comparisons} while only reporting between `r rangeMsqrobFP` false positives (FP) for HeLa proteins. As a result, their FDP ranges between `r rangeMsqrobFDP`%, suggesting appropriate FDR control at the 5% level. The default msTrawler workflow, however, was only able to recover `r msTrawlerTP` TP and reported `r msTrawlerFP` FP, leading to an FDP of `r msTrawlerFDP`%. With our refactored import function this improved to `r msTrawlerTP_fixed` TP, `r msTrawlerFP_fixed` FP and an FDP of `r msTrawlerFDP_fixed`%. The summarisation-based workflows DEqMS and MSstatsTMT, reported `r deqmsTP` and  `r msStatsTmtTP`  TP, respectively, with `r deqmsFP` and `r msStatsTmtFP` FP, resulting in FDPs of `r deqmsFDP`% and `r msStatsTmtFDP`%, respectively, suggesting improper FDR control by DEqMS.

@@

\section{Data Availability}
